\textsc{\wobble: a data-driven method for precision radial velocities}

### 2018-11-28 paper/paper.tex
adding software citations
fixing likelihood notation

Our approach is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals.
In this sense, ours is a true forward model.

@@

In Section \ref{s:methods}, we outline the model and its key underlying assumptions.
We present an open-source implementation of this method in \python and \TF called \wobble.
In Section \ref{s:results}, we demonstrate \wobble's capabilities by applying our method to \HARPS archival data for three target stars: the planet-hosting solar analog 51 Peg, the quiet M dwarf \Mdwarf, and the Hot Jupiter host HD 189733.
We look further into the detailed time-variable telluric spectra inferred from these data in Section \ref{s:tellurics}.
We revisit many of the assumptions underlying \wobble in Section \ref{s:future} with a detailed look at the limitations of the current implementation.
We outline potential ways of adapting \wobble\ for such cases as instruments with absorption cells, intrinsic time variability in the stellar spectrum, and lower-quality data.
Finally, we conclude with a brief summary in Section \ref{s:conclusion}.

@@

We assume that the spectra can be modeled as the product of a finite number of components.
For the cases shown in this work, two components are used: a stellar spectrum which is invariant in shape but may be Doppler-shifted, and a telluric spectrum which is fixed to the observatory rest frame but varies in shape.
We choose to work in log(flux) space so that the data to be modeled are simply a sum of the component spectra.

@@

We model each data column $y_n$ as the sum of the stellar and telluric spectra:
\begin{equation}
y_n = y_{\star, n} + y_{t, n} + noise.
\end{equation}

The stellar spectrum contribution is:
\begin{equation}
y_{\star, n} = P(v_{obs, n}) \mu_{\star},
\end{equation}
where $P$ is an interpolation operator which applies a Doppler shift by velocity $v_n$ and $\mu_{\star}$ is a spectral template with length $M'$ corresponding to the length of an arbitrarily-chosen wavelength grid of control points.
Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
\begin{equation}
 \ln \lambda(v) = \ln \lambda_{0} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
\end{equation}
The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.

@@

The telluric spectrum contribution is:
\begin{equation}
y_{t, n} =  A_n(\mu_{t} + W_{t} z_n).
\end{equation}
In addition to its mean spectrum $\mu_t$, the telluric component also includes a time-dependent term assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
$W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used.
For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications.
The net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.

We evaluate a log-likelihood for each epoch
\begin{equation}
\ln \mathcal{L}_n = \sum_{m} -0.5 (y_{m,n} - y'_{m,n})^T C_{m,n}^{-1} (y_{m,n}-y'_{m,n}),
\end{equation}
comparing our model $y'$ to the data $y$, with $C$ representing the covariance matrix of uncertainties for the data.

@@

L1 normalization adds a term to the log-likelihood that takes the form \todo{(TO DO: change notation to match literature)}:
$$ L1(p, \lambda) = \lambda \sum_{i} | p_{i} | ,$$
where $p$ is the vector of parameters to be normalized (in this case $\mu_{{\star}}$, $\mu_{{t}}$, or $W_{t}$) and $\lambda$ is the regularization amplitude.

Similarly, L2 normalization adds a term of the form:
$$ L2(p, \lambda) = \lambda \sum_{i} p_{i}^2 .$$

@@

\software{\todo{TO DO! --} astropy, numpy, matplotlib, scipy, tensorflow}

>>>>

Our approach is similar to that taken by \citet{Gao2016}, but differs in that we take the stellar spectrum as a latent variable to be optimized rather than iteratively determining it by stacking model residuals.
%In this sense, ours is a true forward model.  -- this content-less (DFM)

@@

In Section \ref{s:methods}, we outline the model and its key underlying assumptions.
We present an open-source implementation of this method in \python and \TF
\citep{Abadi15} called \wobble.
In Section \ref{s:results}, we demonstrate \wobble's capabilities by applying our method to \HARPS archival data for three target stars: the planet-hosting solar analog 51 Peg, the quiet M dwarf \Mdwarf, and the Hot Jupiter host HD 189733.
We look further into the detailed time-variable telluric spectra inferred from these data in Section \ref{s:tellurics}.
We revisit many of the assumptions underlying \wobble in Section \ref{s:future} with a detailed look at the limitations of the current implementation.
We outline potential ways of adapting \wobble\ for such cases as instruments with absorption cells, intrinsic time variability in the stellar spectrum, and lower-quality data.
Finally, we conclude with a brief summary in Section \ref{s:conclusion}.

@@

We assume that the spectra can be modeled as the product of a finite and fixed number of components.
For the cases shown in this work, two components are used: a stellar spectrum which is invariant in shape but may be Doppler-shifted, and a telluric spectrum which is fixed to the observatory rest frame but varies in shape.
We choose to work in log(flux) space so that the data to be modeled are simply a sum of the component spectra.

@@

We model each data column $y_n$ as the sum of the stellar and telluric spectra:
\begin{equation}
y_n = y_{n}^\star + y_{n}^t + noise.
\end{equation}

The stellar spectrum contribution is:
\begin{equation}
y_{n}^\star = P(v_{obs, n}) \mu_{\star},
\end{equation}
where $P$ is an interpolation operator which applies a Doppler shift by velocity $v_n$ and $\mu_{\star}$ is a spectral template with length $M'$ corresponding to the length of an arbitrarily-chosen wavelength grid of control points.
Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
\begin{equation}
 \ln \lambda(v) = \ln \lambda_{0} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
\end{equation}
The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.

@@

The telluric spectrum contribution is:
\begin{equation}
y_{n}^t =  A_n(\mu_{t} + W_{t} z_n).
\end{equation}
In addition to its mean spectrum $\mu_t$, the telluric component also includes a time-dependent term assembled from two variables: $W_t$, a matrix of ``basis vectors'' for the span of telluric spectral variations, is weighted by $z_n$ to form the spectral contribution at epoch $n$.
$W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used.
For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications.
The net telluric spectrum (mean + time-variable components) is weighted by the airmass at the time of observation $A_n$, a known quantity.

We evaluate a log-likelihood for each epoch
\begin{equation}
\ln \mathcal{L}_n = -\frac{1}{2} (y_{n} - y'_{n})^T C_{n}^{-1} (y_{n}-y'_{n}),
\end{equation}
comparing our model $y'$ to the data $y$, with $C$ representing the covariance matrix of uncertainties for the data.

@@

L1 normalization adds a term to the log-likelihood that takes the form \todo{(TO DO: change notation to match literature)}:
$$ L1(p, \lambda) = -\lambda \sum_{i} | p_{i} | ,$$
where $p$ is the vector of parameters to be normalized (in this case $\mu_{{\star}}$, $\mu_{{t}}$, or $W_{t}$) and $\lambda$ is the regularization amplitude.

Similarly, L2 normalization adds a term of the form:
$$ L2(p, \lambda) = -\lambda \sum_{i} p_{i}^2 .$$

@@

\software{%
    astropy \citep{Astropy13, Astropy18},
    numpy \citep{Van-Der-Walt:2011},
    matplotlib \citep{Hunter:2007},
    scipy \citep{Jones:2001},
    tensorflow \citep{Abadi15}
}
